### 2017-03-23 paper/oopsla17-submission/evaluation.tex
say something about experimental setup

\section{Evaluation}
\label{sec:evaluation}

\input{data}

@@

\subsection{Type Error Slice}
\label{sec:type-error-slice}
The \InSlice feature should be highly predictive --- a fix must change
at least one expression in the type-error slice.
%
Thus, our first experiment seeks to quantify the impact of \InSlice by
comparing the accuracy of a linear model on three sets of features.
%
\begin{enumerate}
\item A baseline with only local syntactic features.
\item extends (1) with \InSlice.
\item discards samples from (1) where \InSlice is \emph{disabled}.
\end{enumerate}
%
The key difference between (2) and (3) is that a model for (2) must
\emph{learn} that \InSlice is a strong predictor.
%
In contrast, a model for (3) must only learn about the syntactic
features, the decision to discard samples where \InSlice is disabled has
already been made by a human.
%
This has a few additional advantages: it reduces the search space by a
factor of 7 on average, and it guarantees that any prediction made by
the classifier can fix the type error.
%
We expect that (2) will perform better than (1) as it has more
information, and that (3) will perform better than (2) as it does not
have to learn the importance of \InSlice.

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Feature Set  & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax & 23.6\% & 42.6\% & 56.3\% & 19.6\% \\
    With Slice   & 46.4\% & 65.0\% & 75.4\% & 30.4\% \\
    Only Slice   & 54.9\% & 71.4\% & 82.5\% & 57.6\% \\
    \bottomrule
  \end{tabular}
  \caption{
    Impact of type-error slice on accuracy.
    \ES{TODO: expand caption}
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
\end{table}

@@

\subsection{Contextual Features}
\label{sec:contextual-features}

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    Feature Set                 & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax                & 55.0\% & 71.6\% & 82.6\% & 57.7\% \\
    \midrule
    + Size                      & 55.8\% & 72.8\% & 82.5\% & 57.3\% \\
    + Contextual Syntax         & 59.9\% & 77.7\% & 86.4\% & 63.0\% \\
    + Types                     & 62.2\% & 77.7\% & 85.7\% & 62.2\% \\
    \midrule
    + Size + Contextual Syntax  & 60.2\% & 77.9\% & 86.0\% & 62.5\% \\
    + Size + Types              & 62.0\% & 78.2\% & 85.6\% & 62.3\% \\
    + Contextual Syntax + Types & 63.2\% & 80.3\% & 87.9\% & 65.4\% \\
    \midrule
    + All                       & 62.3\% & 80.0\% & 88.0\% & 65.4\% \\
    \bottomrule
  \end{tabular}
  \caption{
    Impact of contextual features on accuracy.
    \ES{TODO: expand caption}
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
\end{table}

>>>>

\section{Evaluation}
\label{sec:evaluation}
\input{data}

We have implemented our technique for localizing type errors for a
purely functional subset of \ocaml with polymorphic types and functions.
%
We seek to answer three questions in our evaluation:
%
\begin{description}
\item[1. Feature Utility (\autoref{sec:feature-utility})]
  %
  How much do the features described in \autoref{sec:features}
  contribute to our predictions?
\item[2. Blame Accuracy (\autoref{sec:quantitative-evaluation})]
  %
  How often are our predictions \emph{correct}?
  % For how many ill-typed programs can we accurately predict the source
  % of the error?
\item[3. Explaining Predictions (\autoref{sec:qualitative})]
  %
  Can we explain individual predictions (correct or not)?
\end{description}
\ES{may want to swap (1) and (2)..}

@@

We answer our questions on two sets of data gathered from the
undergraduate Programming Languages course at
\if@anonymous
AUTHOR INSTITUTION.
\else
UC San Diego (IRB \#140608).
\fi
%
We recorded each interaction with the \ocaml top-level system over the
course of the first three assignments, which allowed us to capture
ill-typed programs and, crucially, their subsequent fixes.
%
The first dataset comes from the Spring 2014 class (\SPRING), from which
we extracted 2,333 pairs of programs, and the second comes from the Fall
2015 class (\FALL), from which we extracted 2,273 pairs of programs.
%
The extracted programs are relatively small, but they demonstrate a
range of functional programming idioms, \eg higher-order functions and
(polymorphic) algebraic data types.

We identified the fixes for each ill-typed program with an
expression-level diff~\citep{XXXX}.
%
We consider two sources of changes:
%
\begin{enumerate}
\item If an expression has been removed wholesale, \eg if $\eapp{f}{x}$
  is rewritten to $\eapp{g}{x}$, we will consider the expression $f$
  changed, as it has been replaced by $g$.
\item If a new expression has been inserted \emph{around} an existing
  expression, \eg if $\eapp{f}{x}$ is rewritten to
  $\eplus{\eapp{f}{x}}{1}$, we will consider the application expression
  $\eapp{f}{x}$ (but not $f$ or $x$) changed, as the $+$ operator now
  occupies the original location of the application.
\end{enumerate}
%
As the students may have made many, potentially unrelated, changes
between compilations, we discard any program pairs where more than 40\%
of the sub-expressions have changed.
%
We picked 40\% as an estimate of the inflection point where we could
still retain the large majority of program pairs.

\subsection{Feature Utility}
\label{sec:feature-utility}

\subsubsection{Type Error Slice}
\label{sec:type-error-slice}
The \InSlice feature should be highly predictive --- a fix must change
at least one expression in the type-error slice.
%
Thus, our first experiment seeks to quantify the impact of \InSlice by
comparing the accuracy of a linear model on three sets of features.
%
\begin{enumerate}
\item A baseline with only local syntactic features.
\item extends (1) with \InSlice.
\item discards samples from (1) where \InSlice is \emph{disabled}.
\end{enumerate}
%
The key difference between (2) and (3) is that a model for (2) must
\emph{learn} that \InSlice is a strong predictor.
%
In contrast, a model for (3) must only learn about the syntactic
features, the decision to discard samples where \InSlice is disabled has
already been made by a human.
%
This has a few additional advantages: it reduces the search space by a
factor of 7 on average, and it guarantees that any prediction made by
the classifier can fix the type error.
%
We expect that (2) will perform better than (1) as it has more
information, and that (3) will perform better than (2) as it does not
have to learn the importance of \InSlice.

@@

\begin{table}[ht]
  \begin{minipage}{0.49\linewidth}
  \centering
  \linear
  \begin{tabular}{lrrrr}
    \toprule
    Feature Set  & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax & 23.6\% & 42.6\% & 56.3\% & 19.6\% \\
    With Slice   & 46.4\% & 65.0\% & 75.4\% & 30.4\% \\
    Only Slice   & 54.9\% & 71.4\% & 82.5\% & 57.6\% \\
    \bottomrule
  \end{tabular}
  \end{minipage}
  \begin{minipage}{0.49\linewidth}
  \centering
  \hiddenF
  \begin{tabular}{lrrrr}
    \toprule
    Feature Set  & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax & 30.9\% & 47.9\% & 58.6\% & 20.6\% \\
    With Slice   & 54.5\% & 70.5\% & 81.5\% & 34.7\% \\
    Only Slice   & 56.7\% & 72.3\% & 82.9\% & 58.0\% \\
    \bottomrule
  \end{tabular}
  \end{minipage}
  \caption{
    Impact of type-error slice on accuracy.
    \ES{TODO: expand caption}
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
\end{table}

@@

\subsubsection{Contextual Features}
\label{sec:contextual-features}

@@

\begin{table}[ht]
  \begin{minipage}{0.49\linewidth}
  \centering
  \linear
  \begin{tabular}{lrrrr}
    \toprule
    Feature Set                 & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax                & 55.0\% & 71.6\% & 82.6\% & 57.7\% \\
    \midrule
    + Size                      & 55.8\% & 72.8\% & 82.5\% & 57.3\% \\
    + Context                   & 59.9\% & 77.7\% & 86.4\% & 63.0\% \\
    + Types                     & 62.2\% & 77.7\% & 85.7\% & 62.2\% \\
    \midrule
    + Size + Context            & 60.2\% & 77.9\% & 86.0\% & 62.5\% \\
    + Size + Types              & 62.0\% & 78.2\% & 85.6\% & 62.3\% \\
    + Context + Types           & 63.2\% & 80.3\% & 87.9\% & 65.4\% \\
    \midrule
    + All                       & 62.3\% & 80.0\% & 88.0\% & 65.4\% \\
    \bottomrule
  \end{tabular}
  \end{minipage}
  % \begin{minipage}{0.49\linewidth}
  % \centering
  % \hiddenF
  % \begin{tabular}{lrrrr}
  %   \toprule
  %   Feature Set                 & Top-1  & Top-2  & Top-3  & Recall \\
  %   \midrule
  %   Local Syntax                & 56.9\% & 72.2\% & 82.8\% & 57.9\% \\
  %   \midrule
  %   + Size                      & 59.7\% & 74.6\% & 83.0\% & 57.4\% \\
  %   + Context                   & 70.9\% & 83.7\% & 90.4\% & 69.2\% \\
  %   + Types                     & 72.1\% & 84.1\% & 90.3\% & 69.3\% \\
  %   \midrule
  %   + Size + Context            & 69.8\% & 83.5\% & 90.2\% & 68.6\% \\
  %   + Size + Types              & 72.3\% & 84.6\% & 90.3\% & 69.5\% \\
  %   + Context + Types           & 75.5\% & 86.4\% & 91.5\% & 71.7\% \\
  %   \midrule
  %   + All                       & 75.0\% & 86.8\% & 91.9\% & 72.0\% \\
  %   \bottomrule
  % \end{tabular}
  % \end{minipage}
  % \\
  \begin{minipage}{0.49\linewidth}
  \centering
  \hiddenFH
  \begin{tabular}{lrrrr}
    \toprule
    Feature Set                 & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax                & 56.3\% & 72.1\% & 82.8\% & 57.6\% \\
    \midrule
    + Size                      & 60.5\% & 75.0\% & 83.8\% & 57.9\% \\
    + Context                   & 71.4\% & 84.1\% & 90.8\% & 69.5\% \\
    + Types                     & 73.2\% & 84.8\% & 90.2\% & 69.2\% \\
    \midrule
    + Size + Context            & 71.7\% & 83.5\% & 90.7\% & 69.3\% \\
    + Size + Types              & 73.5\% & 85.8\% & 91.4\% & 70.5\% \\
    + Context + Types           & 77.3\% & 87.5\% & 92.4\% & 72.9\% \\
    \midrule
    + All                       & 77.2\% & 87.9\% & 92.7\% & 72.9\% \\
    \bottomrule
  \end{tabular}
  \end{minipage}
  \caption{
    Impact of contextual features on accuracy.
    \ES{TODO: expand caption}
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
\end{table}

### 2017-03-23 paper/oopsla17-submission/learning.tex
say something about experimental setup

\begin{figure}[ht]
\begin{minipage}{0.6\linewidth}
\begin{lstlisting}
  let rec sumList xs =
    match xs with
    | []     -> []
    | hd::tl -> hd + sumList tl
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.3\linewidth}
\begin{lstlisting}[numbers=left]
[]
hd + sumList tl
sumList tl
tl
\end{lstlisting}
\end{minipage}
\caption{An ill-typed program (left) and a selection of its
  sub-expressions (right).}
\label{fig:sumList}
\end{figure}
\begin{table}[ht]
\begin{tabular}{lrrrrrrr}
\toprule
Expression
  & \IsNil & \IsCaseListP & \CountVarP & \ExprSize
  & \HasTypeIntCOne & \HasTypeList & \InSlice \\
\midrule
\lstinline![]!
  & 1 & 1 & 3 & 1 & 0 & 1 & 1 \\
\lstinline!hd + sumList tl!
  & 0 & 1 & 3 & 5 & 1 & 0 & 1 \\
\lstinline!sumList tl!
  & 0 & 0 & 3 & 3 & 0 & 1 & 1 \\
\lstinline!tl!
  & 0 & 0 & 2 & 1 & 0 & 1 & 0 \\
\bottomrule
\end{tabular}
\caption{A selection of the features we would extract from the
  \lstinline!sumList! program in \autoref{fig:sumList}. A feature is
  considered \emph{enabled} if it has a non-zero value, and
  \emph{disabled} otherwise. A ``-P'' suffix indicates that the feature
  describes the parent of the current expression, a ``-C$n$'' suffix
  indicates that the feature describes the $n$-th (left-to-right) child
  of the current expression.  Note that, since we rely on a partial
  typing derivation, we are subject to the well-known traversal bias and
  label the expression \lstinline!sumList tl! as having type
  $\tlist{\cdot}$. The model will have to learn to correct for this
  bias.}
\label{tab:sumList}
\end{table}

@@

We distinguish between matching on a list vs.\ on a pair, as this
affects the typing derivation.
%
We also assume that all pattern matches are well-formed --- \ie all
patterns must match on the same type.
%
Ill-formed match expressions would lead to a type error; however, they
are already effectively localized to the match expression itself.
%
This is not a \emph{fundamental} limitation though, one could easily add
features that specify whether a match \emph{contains} a particular
pattern, and thus have a match expression that enables multiple
features. \ES{This is actually sort of what we do, we just ignore
  programs that have patterns of different types..}

@@

Instead of just describing the immediate context, we could describe
whether a particular syntax element occurs in the neighboring
expressions (or even a count of how many times it occurs).
%
For example, the \CountVarP feature in \autoref{tab:sumList} describes
how many variables are contained in the expression \emph{rooted} at the
current expression's parent.
%
Such fuzzier notions of context may enable increased precision in the
model, but they also introduce opportunities for \emph{overfitting} ---
where the model memorizes particular inputs rather than learning general
patterns.
%
We will investigate (\ES{maybe..}) the impact of these alternatives
in \autoref{sec:evaluation}.

>>>>

\begin{figure}[ht]
\begin{minipage}{0.6\linewidth}
\begin{code}
  let rec sumList xs =
    match xs with
    | []     -> []
    | hd::tl -> hd + sumList tl
\end{code}
\end{minipage}
\begin{minipage}{0.3\linewidth}
\begin{code}[numbers=left]
[]
hd + sumList tl
sumList tl
tl
\end{code}
\end{minipage}
\caption{An ill-typed program (left) and a selection of its
  sub-expressions (right).}
\label{fig:sumList}
\end{figure}
\begin{table}[ht]
\begin{tabular}{lrrrrrr}
\toprule
Expression
  & \IsNil & \IsCaseListP & \ExprSize
  & \HasTypeIntCOne & \HasTypeList & \InSlice \\
\midrule
\lstinline![]!
  & 1 & 1 & 1 & 0 & 1 & 1 \\
\lstinline!hd + sumList tl!
  & 0 & 1 & 5 & 1 & 0 & 1 \\
\lstinline!sumList tl!
  & 0 & 0 & 3 & 0 & 1 & 1 \\
\lstinline!tl!
  & 0 & 0 & 1 & 0 & 1 & 0 \\
\bottomrule
\end{tabular}
\caption{A selection of the features we would extract from the
  \lstinline!sumList! program in \autoref{fig:sumList}. A feature is
  considered \emph{enabled} if it has a non-zero value, and
  \emph{disabled} otherwise. A ``-P'' suffix indicates that the feature
  describes the parent of the current expression, a ``-C$n$'' suffix
  indicates that the feature describes the $n$-th (left-to-right) child
  of the current expression.  Note that, since we rely on a partial
  typing derivation, we are subject to the well-known traversal bias and
  label the expression \lstinline!sumList tl! as having type
  $\tlist{\cdot}$. The model will have to learn to correct for this
  bias.}
\label{tab:sumList}
\end{table}

@@

We distinguish between matching on a list vs.\ on a pair, as this
affects the typing derivation.
%
We also assume that all pattern matches are well-formed --- \ie all
patterns must match on the same type.
%
Ill-formed match expressions would lead to a type error; however, they
are already effectively localized to the match expression itself.
%
This is not a \emph{fundamental} limitation though, one could easily add
features that specify whether a match \emph{contains} a particular
pattern, and thus have a match expression that enables multiple
features.

@@

% Instead of just describing the immediate context, we could describe
% whether a particular syntax element occurs in the neighboring
% expressions (or even a count of how many times it occurs).
% %
% For example, the \CountVarP feature in \autoref{tab:sumList} describes
% how many variables are contained in the expression \emph{rooted} at the
% current expression's parent.
% %
% Such fuzzier notions of context may enable increased precision in the
% model, but they also introduce opportunities for \emph{overfitting} ---
% where the model memorizes particular inputs rather than learning general
% patterns.
% %
% We will investigate (\ES{maybe..}) the impact of these alternatives
% in \autoref{sec:evaluation}.
